---
title: "script_RILI"
output: html_document
date: "2025-04-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
```{r, echo = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("Seurat")
BiocManager::install("EnhancedVolcano")
```

```{r}
library(Seurat)
library(clustree)
library(cluster)
library(ggplot2)
library(dplyr)
library(Matrix)
library(EnhancedVolcano)
```

```{r}
count_matrix <- read.delim("counts_UMI.csv", row.names = 1, check.names = FALSE)
```
create metadata from data provided in the article
```{r}
# Получаем имена колонок
col_names <- colnames(count_matrix)

# Обрезаем имена до первого символа "_"
short_names <- sub("_.*", "", col_names)

metadata <- data.frame(row.names = col_names, shortened = short_names)

# Добавим колонку cell_line
metadata$mouse_line <- ifelse(metadata$shortened %in% c("sfei2", "s2"),
                               "C57BL/6N",
                               ifelse(metadata$shortened %in% c("s3", "s4"),
                                      "C3H/HeN", NA))

# Добавим колонку EC
metadata$EC <- ifelse(metadata$shortened %in% c("sfei2", "s3"),
                        "control",
                        ifelse(metadata$shortened %in% c("s2", "s4"),
                               "experiment", NA))

# Удалим колонку shortened
metadata$shortened <- NULL

write.csv(metadata, "metadata.csv")
```

```{r}
metadata <- read.csv("metadata.csv")
```

```{r}
# Create a Seurat object
seurat_object <- CreateSeuratObject(counts = count_matrix, meta.data = metadata, min.cells = 3, min.features = 200)
```
```{r}
#data cleaning - mtRNA
#seurat_object[["percent.mt"]] <- PercentageFeatureSet(seurat_object, pattern = "^MT-")

#clean_plot <- VlnPlot(seurat_object, features = c("nFeature_RNA", "nCount_RNA",'percent.mt'), ncol = 3)
#clean_plot

#according to this plot, we see that all the cells have percent.mt < 5 (or mt genes are not presentedd in the data table)

# FeatureScatter is typically used to visualize feature-feature relationships, but can be used for anything calculated by the object, i.e. columns in object metadata, PC scores etc.
#FeatureScatter(seurat_object, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

#according to this plot, we remove outlets
#seurat_object <- subset(seurat_object, subset = nFeature_RNA > 200 & nFeature_RNA < 6000 & percent.mt < 5)
seurat_object <- subset(seurat_object, subset = nFeature_RNA > 200 & nFeature_RNA < 6000)
```
```{r, echo=TRUE}
BiocManager::install("SingleR")
BiocManager::install("BiocGenerics")
BiocManager::install("Biobase")
BiocManager::install("BiocNeighbors")
BiocManager::install("BiocParallel")
BiocManager::install("BiocSingular")
install.packages("bit64")
install.packages("blob")
BiocManager::install("Biostrings")
BiocManager::install("celldex")
BiocManager::install("SingleCellExperiment")
```

# Cell annotation
```{r, echo=TRUE}
library(SingleR)
library(celldex)
library(SingleCellExperiment)

ref <- celldex::MouseRNAseqData()
#ref <- celldex::ImmGenData()

pred <- SingleR(test = seurat_object[["RNA"]]$counts, ref = ref, labels = ref$label.main)

seurat_object$cell_type <- pred$labels
```

#divide into two subsets according to mouse line
```{r}
seuratC57BL <- subset(seurat_object, subset = mouse_line == 'C57BL/6N')
seuratC3H <- subset(seurat_object, subset = mouse_line == 'C3H/HeN')

```

#Standard workflow with SCTransform
```{r}
#install.packages("future")
library(future)

#Main problem: SCTransform() function uses large objects (e.g. FUN, umi_bin, data_step1) and they are too heavy to be passed to parallel session with current limit.
# Increase global variables size limit 

options(future.globals.maxSize = 2 * 1024^3)  # 2 GB
seuratC57BL <- SCTransform(seuratC57BL, verbose = FALSE)
seuratC3H <- SCTransform(seuratC3H, verbose = FALSE)
```

```{r}
#look at most variable genes in both subsets
top10C57BL <- head(VariableFeatures(seuratC57BL), 10)
plotC57BL_vf <- LabelPoints(plot = VariableFeaturePlot(seuratC57BL), points = top10C57BL, repel = TRUE)

top10C3H <- head(VariableFeatures(seuratC3H), 10)
plotC3H_vf <- LabelPoints(plot = VariableFeaturePlot(seuratC3H), points = top10C3H, repel = TRUE)

plotC57BL_vf
plotC3H_vf
```


```{r}
seuratC57BL <- RunPCA(seuratC57BL, features = VariableFeatures(object = seuratC57BL))
seuratC57BL <- RunUMAP(seuratC57BL, dims = 1:30)
seuratC57BL <- FindNeighbors(seuratC57BL, dims = 1:30)
seuratC57BL <- FindClusters(seuratC57BL, resolution = 0.5)

#DimPlot(seuratC57BL, reduction = "umap", group.by = "EC", repel = TRUE)
#DimPlot(seuratC57BL, reduction = "umap", label = TRUE, label.box = TRUE, repel = TRUE) + NoLegend()
```

```{r}
seuratC3H <- RunPCA(seuratC3H, features = VariableFeatures(object = seuratC3H))
seuratC3H <- RunUMAP(seuratC3H, dims = 1:30)
seuratC3H <- FindNeighbors(seuratC3H, dims = 1:30)
seuratC3H <- FindClusters(seuratC3H, resolution = 0.5)

#DimPlot(seuratC3H, reduction = "umap", group.by = "EC", repel = TRUE)
#DimPlot(seuratC3H, reduction = "umap", label = TRUE, label.box = TRUE, repel = TRUE) + NoLegend()
```

```{r}
DimHeatmap(seuratC57BL, dims = 1:5, cells = 500, balanced = TRUE)
```

```{r}
# Convenience functions
fig_path <- ""
SaveFigure <- function(plots, name, type = "png", width, height, res){
  if(type == "png") {
    png(paste0(fig_path, name, ".", type),
      width = width, height = height, units = "in", res = 200)
  } else {
    pdf(paste0(fig_path, name, ".", type),
      width = width, height = height)
  }
  print(plots)
  dev.off()
}

SaveObject <- function(object, name){
  saveRDS(object, paste0(data_path, name, ".RDS"))
}

data_path <- ""
ReadObject <- function(name){
  readRDS(paste0(data_path, name, ".RDS"))
}
```

#find markers for clusters and create heatmap
```{r}
seuratC57BL.markers <- FindAllMarkers(seuratC57BL, only.pos = TRUE)
seuratC57BL.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10C57BL
#marker_map_C57BL <- DoHeatmap(seuratC57BL, features = top10C57BL$gene) + NoLegend()
#SaveFigure(marker_map_C57BL, "markers_C57BL_save", width = 30, height = 30, res = 200)
```

```{r}
seuratC3H.markers <- FindAllMarkers(seuratC3H, only.pos = TRUE)
seuratC3H.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10C3H
#marker_map_C3H <- DoHeatmap(seuratC3H, features = top10C3H$gene) + NoLegend()
#SaveFigure(marker_map_C3H, "markers_C3H_save", width = 30, height = 30, res = 200)
```

#cell type plot
```{r}
DimPlot(seuratC57BL, group.by = "cell_type",
        label = TRUE, 
        repel = TRUE, 
        label.size = 5) + 
  NoLegend()

DimPlot(seuratC3H, group.by = "cell_type",
        label = TRUE, 
        repel = TRUE, 
        label.size = 5) + 
  NoLegend()
```

#mapping clusters with cell types
#for seuratC57BL
```{r}
# Table of cell types per cluster
table <- table(seuratC57BL$seurat_clusters, seuratC57BL$cell_type)

# Convert to a data frame for easier manipulation
cluster_annotations <- as.data.frame.matrix(table)

ec_labels <- as.data.frame.matrix(table(seuratC57BL$seurat_clusters, seuratC57BL$EC))

# Get the top annotation per cluster
top_annotation <- apply(cluster_annotations, 1, function(x) names(x)[which.max(x)])

ec_annotation <- apply(ec_labels, 1, function(x) names(x)[which.max(x)])

# View cluster-to-cell type mapping
cluster_map_C57 <- data.frame(
  Cluster = rownames(cluster_annotations),
  Cell_Type = top_annotation,
  EC = ec_annotation
)

#write.csv(cluster_map_C57, "cluster_map_C57")

# Create a new column with cluster labels mapped to cell types
seuratC57BL$ClusterAnnotation <- cluster_map_C57$Cell_Type[match(seuratC57BL$seurat_clusters, cluster_map_C57$Cluster)]

seuratC57BL$ECAnnotation <- cluster_map_C57$EC[match(seuratC57BL$seurat_clusters, cluster_map_C57$Cluster)]

# Colored by cluster-dominant cell type mapping
#DimPlot(seuratC57BL, group.by = "ClusterAnnotation", label = TRUE, repel = TRUE) + 
  #ggtitle("Clusters Mapped to Dominant Cell Types") + NoLegend()
```

```{r}
dotC57 <- DotPlot(
  seuratC57BL,
  features = unique(top10C57BL$gene),  # or pick genes of interest
  group.by = "ClusterAnnotation",
  split.by = "ECAnnotation"
) + RotatedAxis() + NoLegend()
SaveFigure(dotC57, "dotC57", width = 40, height = 20, res = 200)
```

#for seuratC3H
```{r}
# Table of cell types per cluster
table <- table(seuratC3H$seurat_clusters, seuratC3H$cell_type)

# Convert to a data frame for easier manipulation
cluster_annotations <- as.data.frame.matrix(table)

ec_labels <- as.data.frame.matrix(table(seuratC3H$seurat_clusters, seuratC3H$EC))

# Get the top annotation per cluster
top_annotation <- apply(cluster_annotations, 1, function(x) names(x)[which.max(x)])

ec_annotation <- apply(ec_labels, 1, function(x) names(x)[which.max(x)])

# View cluster-to-cell type mapping
cluster_map_C3H <- data.frame(
  Cluster = rownames(cluster_annotations),
  Cell_Type = top_annotation,
  EC = ec_annotation
)

write.csv(cluster_map_C3H, "cluster_map_C3H")

# Create a new column with cluster labels mapped to cell types
seuratC3H$ClusterAnnotation <- cluster_map_C3H$Cell_Type[match(seuratC3H$seurat_clusters, cluster_map_C3H$Cluster)]

seuratC3H$ECAnnotation <- cluster_map_C3H$EC[match(seuratC3H$seurat_clusters, cluster_map_C3H$Cluster)]

# Colored by cluster-dominant cell type mapping
#DimPlot(seuratC3H, group.by = "ClusterAnnotation", label = TRUE, repel = TRUE) + 
  #ggtitle("Clusters Mapped to Dominant Cell Types") + NoLegend()
```

```{r}
dotC3H <- DotPlot(
  seuratC3H,
  features = unique(top10C3H$gene),  # or pick genes of interest
  group.by = "ClusterAnnotation",
  split.by = "ECAnnotation"
) + RotatedAxis() + NoLegend()
SaveFigure(dotC3H, "dotC3H", width = 40, height = 20, res = 200)
```

#manual cluster mapping
#for C57BL
```{r}
Idents(seuratC57BL) <- seuratC57BL$seurat_clusters
#macrophages
VlnPlot(seuratC57BL, features = c("Cd44", "Lyz2", "C1qa"))
#Cd44 - 0,2; Lyz2 - 0,2,5,7, C1qa - 5

#monocytes
VlnPlot(seuratC57BL, features = c("Ly6c2", "Ms4a6c", "Csf1r", "F13a1" ))
#Ly6c2 - 7,9, Ms4a6c - 5,7, Csf1r - 5,7, F13a1 - 5,7

#granulocytes
VlnPlot(seuratC57BL, features = c("Mpo"))
# Mpo - 1,5,7, 24

#T and NK cells
VlnPlot(seuratC57BL, features = c("Cd3g", "Gzma", "Nkg7", "Klre1"))
#Cd3g 3, 9,10; Gzma - 9,10,22, Nkg7 - 9,10,16,22, Klre1 - 16

#B cells
VlnPlot(seuratC57BL, features = c("Cd79a", "Cd79b", "Ighm", "Igkc"))
#Cd79a - 4, Cd79b - 4, Ighm -, Igkc -

#dendritic cells
VlnPlot(seuratC57BL, features = c("Ccl17", "H2-Eb1", "Cd209a", "Cd74"))
#Ccl17 - 18, H2-Eb1 - 18,13,20, Cd209a - 18, Cd74 - 5,7,18,20,13

#endothelial cells
VlnPlot(seuratC57BL, features = c("Cldn5", "Kdr", "Cdh5", "Pecam1", "Eng"))
#Cldn5 - 6,12,13,21, Kdr - 6,12,13,21,23, Cdh5 -  6,12,13,21, Pecam1 -  6,12,13,21, Eng -  6,12,13,21

#fibroblasts
VlnPlot(seuratC57BL, features = c("Dcn", "Gsn", "Col1a2", "Col3a1", "Col1a1", "Mgp", "Gpx3"))
#Dcn - 1,11,23, Gsn - 1,8,11,14, Col1a2 - 1,8,11, Col3a1 - 1,8,11, Col1a1 - 1,8,11, Mgp - 1,8,11, 23, Gpx3 - 1,11,12,23

#smooth muscle cells
VlnPlot(seuratC57BL, features = c("Acta2", "Myl9", "Mustn1"))
#Acta2 - 8, Myl9 - 8, Mustn1 - 8

#epithelial cells
VlnPlot(seuratC57BL, features = c("Krt18", "Krt8", "Foxj1", "Sntn"))
#Krt18 - 14,19,20,23,25, Krt8 - 14,19,20,23,25, Foxj1 - 14, Sntn - 14

#0 - macrophages
#1 - fibroblasts
#2 - macrophages
#3 - 
#4 - B cells
#5 - monocytes
#6 - endothelial cells
#7 - monocytes
#8 - smooth muscle cells
#9 - T and NK cells
#10 - T and NK cells
#11 - fibroblasts
#12 - endothelial cells
#13 - endothelial cells, dendritic cells
#14 - epithelial cells
#15 - 
#16 - T and NK cells
#17 -
#18 - dendritic cells
#19 - epithelial cells
#20 - epithelial cells, dendritic cells
#21 - endothelial cells
#22 - T and NK cells
#23 - fibroblasts
#24 - granulocytes
#25 - epithelial cells

map <- c(
  "macrophages",
  "fibroblasts",
  "macrophages",
  "_",
  "B cells",
  "monocytes",
  "endothelial cells",
  "monocytes",
  "smooth muscle cells",
  "T and NK cells",
  "T and NK cells",
  "fibroblasts",
  "endothelial cells",
  "endothelial cells, dendritic cells",
  "epithelial cells",
  "_",
  "T and NK cells",
  "_",
  "dendritic cells",
  "epithelial cells",
  "epithelial cells, dendritic cells",
  "endothelial cells",
  "T and NK cells",
  "fibroblasts",
  "granulocytes",
  "epithelial cells"
)
manual_map <- data.frame(cluster = levels(seuratC57BL), map = map)

seuratC57BL$ManualCluster <- manual_map$map[match(seuratC57BL$seurat_clusters, manual_map$cluster)]

# Colored by cluster-dominant cell type mapping
DimPlot(seuratC57BL, group.by = "ManualCluster", label = TRUE, repel = TRUE) + 
  ggtitle("Clusters Mapped to Dominant Cell Types") + NoLegend()
```


# Apply the mapping
#seuratC57BL$ManualCluster <- manual_map[as.character(seuratC57BL$seurat_clusters)]


manual_map <- c(
  "0" =  "macrophages",
"1" = "fibroblasts",
"2" = "macrophages",
"3" = "",
"4" = "B cells",
"5" = "monocytes",
"6" = "endothelial cells",
"7" = "monocytes",
"8" = "smooth muscle cells",
"9" = "T and NK cells",
"10" = "T and NK cells",
"11" = "fibroblasts",
"12" = "endothelial cells",
"13" = "endothelial cells, dendritic cells",
"14" = "epithelial cells",
"15" = "",
"16" = "T and NK cells",
"17" = "",
"18" = "dendritic cells",
"19" = "epithelial cells",
"20" = "epithelial cells, dendritic cells",
"21" = "endothelial cells",
"22" = "T and NK cells",
"23" = "fibroblasts",
"24" = "granulocytes",
"25" = "epithelial cells"
)

#Heatmap for clusters with EC and cell type
#for C57BL
```{r}
# Combine experimental condition and cell type into one grouping variable
seuratC57BL$HeatmapGroup <- paste(seuratC57BL$ECAnnotation, seuratC57BL$ClusterAnnotation,   seuratC57BL$seurat_clusters, sep = "_")

# Set it as the active identity
Idents(seuratC57BL) <- "HeatmapGroup"

heatC57 <- DoHeatmap(seuratC57BL, features = top10C57BL$gene, angle = 90) + NoLegend()

SaveFigure(heatC57, "heatC57", width = 30, height = 40, res = 200)
```

#for C3H
```{r}
# Combine experimental condition and cell type into one grouping variable
seuratC3H$HeatmapGroup <- paste(seuratC3H$ECAnnotation, seuratC3H$ClusterAnnotation,   seuratC3H$seurat_clusters, sep = "_")

# Set it as the active identity
Idents(seuratC3H) <- "HeatmapGroup"

heatC3H <- DoHeatmap(seuratC3H, features = top10C3H$gene, angle = 90) + NoLegend()

SaveFigure(heatC3H, "heatC3H", width = 30, height = 40, res = 200)
```

#DE for different cell types between experiment and control
#for C57BL
```{r}
Idents(seuratC57BL) <- "cell_type"
celltypes <- levels(seuratC57BL)
# List to store DE results
de_list <- list()

for (celltype in celltypes) {
  cat("Running DE for:", celltype, "\n")
  
  # Subset to one cell type
  seurat_obj <- subset(seuratC57BL, idents = celltype)
  
  # Check counts per condition
  counts <- table(seurat_obj$EC)

  if (length(counts) < 2 || any(counts <= 3)) {
    cat("Skipping", celltype, "- not enough cells in one or both groups\n")
    next
  }

  Idents(seurat_obj) <- "EC"
  
  # Run DE: Experiment vs Control
  #avg_logFC: log fold-chage of the average expression between the two groups. Positive values indicate that the gene is more highly expressed in the first group
  de <- FindMarkers(
    seurat_obj,
    ident.1 = "experiment",
    ident.2 = "control",
    logfc.threshold = 0.1,
    min.pct = 0.1,
    return.thresh = 1
  )
  
  # Add gene names and cell type info
  de$gene <- rownames(de)
  de$celltype <- celltype
  
  # Store
  de_list[[celltype]] <- de
}
# Combine into one big table
all_de <- bind_rows(de_list)

# Plot function
plot_volcano <- function(df, celltype, top_n = 10) {

  EnhancedVolcano(df,
                lab = rownames(df),
                x = "avg_log2FC",
                y = "p_val_adj",
                title = paste("Volcano Plot:", celltype),
                pCutoff = 0.05,
                FCcutoff = 1.2,
                axisLabSize = 8,
                titleLabSize = 8,
                subtitleLabSize = 6,
                captionLabSize = 6,
                legendLabSize = 6,
                legendIconSize = 3) #+
          #ggplot2::coord_cartesian(xlim=c(-1, 1), ylim = c(0,3)) + 
          #ggplot2::scale_x_continuous(breaks=seq(-1,1, 1)) + 
          #ggplot2::scale_y_continuous(breaks=seq(0,3, 1))
}
  
  


# Plot volcano for each cell type
for (celltype in celltypes) {
 if (!is.null(de_list[[celltype]])) {
    volcano <- plot_volcano(de_list[[celltype]], celltype)
    print(volcano)
  } else {
    cat("Skipped:", celltype, "- no DE data\n")
  }
}
```

#for C3H
```{r}
Idents(seuratC3H) <- "cell_type"
celltypes_C3H <- levels(seuratC3H)
# List to store DE results
de_list_C3H <- list()

for (celltype in celltypes_C3H) {
  cat("Running DE for:", celltype, "\n")
  
  # Subset to one cell type
  seurat_obj <- subset(seuratC3H, idents = celltype)
  
  # Check counts per condition
  counts <- table(seurat_obj$EC)

  if (length(counts) < 2 || any(counts <= 3)) {
    cat("Skipping", celltype, "- not enough cells in one or both groups\n")
    next
  }

  Idents(seurat_obj) <- "EC"
  
  # Run DE: Experiment vs Control
  #avg_logFC: log fold-chage of the average expression between the two groups. Positive values indicate that the gene is more highly expressed in the first group
  de <- FindMarkers(
    seurat_obj,
    ident.1 = "experiment",
    ident.2 = "control",
    logfc.threshold = 0.1,
    min.pct = 0.1,
    return.thresh = 1
  )
  
  # Add gene names and cell type info
  de$gene <- rownames(de)
  de$celltype <- celltype
  
  # Store
  de_list_C3H[[celltype]] <- de
}
# Combine into one big table
all_de_C3H <- bind_rows(de_list_C3H)

# Plot function
plot_volcano <- function(df, celltype, top_n = 10) {

  EnhancedVolcano(df,
                lab = rownames(df),
                x = "avg_log2FC",
                y = "p_val_adj",
                title = paste("Volcano Plot:", celltype),
                pCutoff = 0.05,
                FCcutoff = 1.2,
                axisLabSize = 8,
                titleLabSize = 8,
                subtitleLabSize = 6,
                captionLabSize = 6,
                legendLabSize = 6,
                legendIconSize = 3) #+
          #ggplot2::coord_cartesian(xlim=c(-1, 1), ylim = c(0,3)) + 
          #ggplot2::scale_x_continuous(breaks=seq(-1,1, 1)) + 
          #ggplot2::scale_y_continuous(breaks=seq(0,3, 1))
}
  
  


# Plot volcano for each cell type
for (celltype in celltypes_C3H) {
 if (!is.null(de_list_C3H[[celltype]])) {
    volcano <- plot_volcano(de_list_C3H[[celltype]], celltype)
    print(volcano)
  } else {
    cat("Skipped:", celltype, "- no DE data\n")
  }
}
```


```{r}
BiocManager::install("clusterProfiler")
BiocManager::install("org.Mm.eg.db")
BiocManager::install("enrichplot")
BiocManager::install("aplot")
BiocManager::install("ape")
```

#functional analysis
#for C57BL
```{r}
library(clusterProfiler)
library(org.Mm.eg.db)
library(enrichplot)

go_results_list <- list()

for (celltype in celltypes) {
  if (!is.null(de_list[[celltype]])) {
    genes <- de_list[[celltype]] %>%
  filter(avg_log2FC > 1, p_val_adj < 0.05) %>% row.names()
  
  entrez_genes <- bitr(genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Mm.eg.db)
  
  if (!is.null(entrez_genes) && nrow(entrez_genes) > 0) {
    go_res <- enrichGO(
      gene = entrez_genes$ENTREZID,
      OrgDb = org.Mm.eg.db,
      keyType = "ENTREZID",
      ont = "BP",  # Biological Process
      pAdjustMethod = "BH",
      pvalueCutoff = 0.05,
      qvalueCutoff = 0.2
    )
    
    if (!is.null(go_res) && nrow(go_res) > 0) {
      go_results_list[[celltype]] <- go_res
      
    }
  } 
  }
}

for (cl in names(go_results_list)) {
  go_res <- go_results_list[[cl]]
  if (!is.null(go_res) && nrow(go_res) > 0) {
    print(
      dotplot(go_res, showCategory = 10) + 
      ggtitle(paste("GO dotplot ", cl))
    )
  }
}


for (cl in names(go_results_list)) {
  go_res <- go_results_list[[cl]]
  if (!is.null(go_res) && nrow(go_res) > 0) {
    go_res_sim <- pairwise_termsim(go_res)
    print(
      emapplot(go_res_sim, showCategory = 10) + 
      ggtitle(paste("GO emapplot", cl))
    )
  }
}
```

#for C3H
```{r}
go_results_list_C3H <- list()

for (celltype in celltypes_C3H) {
  if (!is.null(de_list_C3H[[celltype]])) {
    genes <- de_list_C3H[[celltype]] %>%
  filter(avg_log2FC > 1, p_val_adj < 0.05) %>% row.names()
  
  entrez_genes <- bitr(genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Mm.eg.db)
  
  if (!is.null(entrez_genes) && nrow(entrez_genes) > 0) {
    go_res <- enrichGO(
      gene = entrez_genes$ENTREZID,
      OrgDb = org.Mm.eg.db,
      keyType = "ENTREZID",
      ont = "BP",  # Biological Process
      pAdjustMethod = "BH",
      pvalueCutoff = 0.05,
      qvalueCutoff = 0.2
    )
    
    if (!is.null(go_res) && nrow(go_res) > 0) {
      go_results_list_C3H[[celltype]] <- go_res
      
    }
  } 
  }
}

for (cl in names(go_results_list_C3H)) {
  go_res <- go_results_list_C3H[[cl]]
  if (!is.null(go_res) && nrow(go_res) > 0) {
    print(
      dotplot(go_res, showCategory = 10) + 
      ggtitle(paste("GO dotplot ", cl))
    )
  }
}


for (cl in names(go_results_list_C3H)) {
  go_res <- go_results_list_C3H[[cl]]
  if (!is.null(go_res) && nrow(go_res) > 0) {
    go_res_sim <- pairwise_termsim(go_res)
    print(
      emapplot(go_res_sim, showCategory = 10) + 
      ggtitle(paste("GO emapplot", cl))
    )
  }
}
```

#clustering inside separate cell types
```{r}
Idents(seurat_object) <- "cell_type"
celltypes_total <- levels(seurat_object)

for (type in celltypes_total) {
  cell_obj <- subset(seurat_object, idents = type)
  
  cell_obj <- SCTransform(cell_obj, verbose = FALSE)
    cell_obj <- RunPCA(cell_obj, features = VariableFeatures(object = cell_obj))
  cell_obj <- RunUMAP(cell_obj, dims = 1:30)
  cell_obj <- FindNeighbors(cell_obj, dims = 1:30)
  cell_obj <- FindClusters(cell_obj, resolution = 0.5)
  
  cell_obj.markers <- FindAllMarkers(cell_obj, only.pos = TRUE)
  name <- paste(type, "_obj")
  SaveObject(cell_obj, name)
}

for (type in celltypes_total) {
  cell_obj <- ReadObject(paste(type, "_obj"))
  
  cell_obj.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10cell_obj
  
    ec_labels <- as.data.frame.matrix(table(cell_obj$seurat_clusters, cell_obj$EC))
  
  ec_annotation <- apply(ec_labels, 1, function(x) names(x)[which.max(x)])
  
  # View cluster-to-cell type mapping
  cluster_map <- data.frame(
    Cluster = rownames(ec_labels),
    EC = ec_annotation
  )
  
  # Create a new column with cluster labels mapped to cell types
  cell_obj$ECAnnotation <- cluster_map$EC[match(cell_obj$seurat_clusters, cluster_map$Cluster)]
  
  cell_obj$HeatmapGroup <- paste(cell_obj$ECAnnotation, cell_obj$seurat_clusters, sep = "_")

  # Set it as the active identity
  Idents(cell_obj) <- "HeatmapGroup"
  
  heat <- DoHeatmap(cell_obj, features = top10cell_obj$gene, angle = 90) + NoLegend() + ggtitle(paste("heatmap ", type))
  
  name = paste("heatmap_", type)
  
  SaveFigure(heat, name, width = 25, height = 25, res = 200)
} 
```
#DE for different cell types (whole dataset)
```{r}
total_de_list <- list()
for (type in celltypes_total) {
  cell_obj <- try(ReadObject(paste(type, "_obj")))
  
  if (inherits(cell_obj, "try-error")) {
    cat("error reading object", type)
    next
  }
  # Check counts per condition
  counts <- table(cell_obj$EC)

  if (length(counts) < 2 || any(counts <= 3)) {
    cat("Skipping", type, "- not enough cells in one or both groups\n")
    next
  }
  
  Idents(cell_obj) <- "EC"
  
  de <- FindMarkers(
    cell_obj,
    ident.1 = "experiment",
    ident.2 = "control",
    logfc.threshold = 0.1,
    min.pct = 0.1,
    return.thresh = 1
  )
  
  # Add gene names and cell type info
  de$gene <- rownames(de)
  de$celltype <- type
  
  # Store
  total_de_list[[type]] <- de
}

for (type in celltypes_total) {
  if (!is.null(total_de_list[[type]])) {
    df <- total_de_list[[type]]
  volc <- EnhancedVolcano(df,
                lab = rownames(df),
                x = "avg_log2FC",
                y = "p_val_adj",
                title = paste("Volcano Plot:", type),
                pCutoff = 0.05,
                FCcutoff = 1.2,
                axisLabSize = 8,
                titleLabSize = 8,
                subtitleLabSize = 6,
                captionLabSize = 6,
                legendLabSize = 6,
                legendIconSize = 3) #+
          #ggplot2::coord_cartesian(xlim=c(-1, 1), ylim = c(0,3)) + 
          #ggplot2::scale_x_continuous(breaks=seq(-1,1, 1)) + 
          #ggplot2::scale_y_continuous(breaks=seq(0,3, 1))
  
  name = paste("volc_total_", type)
  SaveFigure(volc, name, width = 25, height = 25, res = 200)
  }
}
```

#ratio of different cell types in control and after radiation
```{r}
EC <- unique(seurat_object$EC)
mouse_line <- unique(seurat_object$mouse_line)

Idents(seurat_object) <- "EC"

for (c in EC) {
  # Subset to one cell type
  obj1 <- subset(seurat_object, idents = c)
  Idents(obj1) <- "mouse_line"
  levels(obj1)
  
  for (l in mouse_line) {
    obj2 <- subset(obj1, idents = l)
    Idents(obj2) <- "cell_type"
    celltypes <- levels(obj2)
    counts <- as.data.frame(table(obj2$cell_type))
    
    counts$percent <- round(counts$Freq / sum(counts$Freq) * 100, 1)
    counts <- counts %>% filter(percent > 1)
    colnames(counts) <- c("celltype", "Freq", "percent")

    plot <-ggplot(counts, aes(x = "", y = Freq, fill = celltype)) +
      geom_bar(stat = "identity", width = 1, color = "white") +
      coord_polar("y") +
      geom_text(aes(label = percent), position = position_stack(vjust = 0.5)) +
      theme_void() +
      labs(title = paste(l, c)) +
      theme(legend.position = "right")
    print(plot)
      }
}


```



#alternative chanks, which are not shown in presentation
#additional look at upregulated and downregulated genes
```{r}
for (type in celltypes_total) {
  if (!is.null(total_de_list[[type]])) {
    df <- total_de_list[[type]]
    df %>% arrange(desc(avg_log2FC))
    print(type)
    print(row.names(head(df, 10)))
    print(row.names(tail(df, 10)))
  }
}
```
#pipeline without SCTransform
```{r}
seurat_object <- FindVariableFeatures(seurat_object, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(seurat_object), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(seurat_object)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2

#Next, we apply a linear transformation (‘scaling’) that is a standard pre-processing step prior to dimensional reduction techniques like PCA. The ScaleData() function:

#Shifts the expression of each gene, so that the mean expression across cells is 0
#Scales the expression of each gene, so that the variance across cells is 1
#This step gives equal weight in downstream analyses, so that highly-expressed genes do not dominate
#The results of this are stored in pbmc[["RNA"]]$scale.data
#By default, only variable features are scaled.
#You can specify the features argument to scale additional features
#all.genes <- rownames(pbmc)
#pbmc <- ScaleData(pbmc, features = all.genes)
seurat_object <- ScaleData(seurat_object)
```


#collect top10 markers for each claster in dataframe
```{r}
cluster.markers <- FindMarkers(seuratC57BL, ident.1 = 0)
top10markers <- head(cluster.markers, n = 10)
dfC57BL<-data.frame(1, row.names(top10markers))
names(dfC57BL)<-c("cluster","genes")
library(tidyverse)
for (x in 1:22) {
  cluster.markers <- FindMarkers(seuratC57BL, ident.1 = x)
  top10markers <- head(cluster.markers, n = 10)
  dfC57BL %>% add_row(cluster = x, genes = row.names(top10markers))
}
write.csv(dfC57BL, "dfC57BL.csv")
```

```{r}
cluster.markers <- FindMarkers(seuratC3H, ident.1 = 0)
top10markers <- head(cluster.markers, n = 10)
dfC3H<-data.frame(1, row.names(top10markers))
names(dfC3H)<-c("cluster","genes")
for (x in 1:23) {
  cluster.markers <- FindMarkers(seuratC3H, ident.1 = x)
  top10markers <- head(cluster.markers, n = 10)
  dfC3H %>% add_row(cluster = x, genes = row.names(top10markers))
}
write.csv(dfC3H, "dfC3H.csv")
```

#attempt to create complex heatmap
```{r}
#install.packages("circlize")
#BiocManager::install("ComplexHeatmap")

library(ComplexHeatmap)
library(circlize)

# Get cell-level metadata
meta <- seuratC57BL@meta.data
meta$cell_id <- rownames(meta)

# Prepare sort key
meta$ECAnnotation <- factor(meta$ECAnnotation)  # ensure consistent order
meta$ClusterAnnotation <- factor(meta$ClusterAnnotation)
meta$seurat_clusters <- factor(meta$seurat_clusters, levels = sort(as.numeric(unique(meta$seurat_clusters))))

# Sort metadata by EC → Cell Type → Cluster
meta_sorted <- meta[order(meta$ECAnnotation, meta$ClusterAnnotation, meta$seurat_clusters), ]

# Use scaled data (used in DoHeatmap)
expr_matrix <- as.matrix(
  GetAssayData(
    seuratC57BL, 
    layer = "data", 
    assay = "SCT")[
    top10C57BL$gene, 
    ]
  )

expr_matrix <- heatmap_matrix[, meta_sorted$cell_id]

## scale the rows
scaled_matrix<- t(scale(t(expr_matrix)))

# Keep only top marker genes
#top_genes <- unique(top10$gene)
#heatmap_matrix <- scaled_matrix[top_genes, ]

# Make sure cells (columns) are in the same order as the matrix
cells_order <- colnames(heatmap_matrix)

# Create annotations
cluster_ann <- seuratC57BL$seurat_clusters[cells_order]
celltype_ann <- seuratC57BL$ClusterAnnotation[cells_order]
condition_ann <- seuratC57BL$ECAnnotation[cells_order]

# Combine into a data frame
column_annotations <- data.frame(
  Cluster = cluster_ann,
  CellType = celltype_ann,
  Condition = condition_ann
)

# Optional: set colors
cluster_colors <- rainbow(length(unique(cluster_ann)))
names(cluster_colors) <- unique(cluster_ann)

celltype_colors <- RColorBrewer::brewer.pal(length(unique(celltype_ann)), "Set3")
names(celltype_colors) <- unique(celltype_ann)

condition_colors <- c("control" = "skyblue", "experiment" = "tomato")

# Recreate the Seurat expression scale
palette_expression_level = circlize::colorRamp2(
  c(min(expr_matrix), median(expr_matrix), min(2.5, max(expr_matrix))),
  c("#FF00FF", "#000000", "#FFFF00"))


# Create ComplexHeatmap annotations
col_ha <- HeatmapAnnotation(
  Condition = condition_ann,
  CellType = celltype_ann,
  Cluster = cluster_ann,
  col = list(
    Cluster = cluster_colors,
    CellType = celltype_colors,
    Condition = condition_colors
  ),
  annotation_name_side = "left"
)

Heatmap(
  scaled_matrix,
  name = "Expression",
  top_annotation = col_ha,
  show_column_names = FALSE,
  show_row_names = TRUE,
  cluster_columns = FALSE,
  cluster_rows = FALSE,
  # Color scale for expression
  col = palette_expression_level,
  # Otherwise you get random vertical white lines where there should not be any
  use_raster = FALSE
)
```




